---
title: "bayesion model"
author: "Krithika Natarajan"
date: "2026-04-09"
output: pdf_document
---



```{r setup, include=FALSE}
install.packages(c('tidyr', 'dplyr', 'brms', 'ggplot2'))
library(tidyr)
library(dplyr)
library(ggplot2)
library(brms)
```



## Reading in the Data

```{r}
df <- read.csv('boston_cleaned_data.csv', colClasses = c(ZipCode = 'character'))
 
meta_cols <- c('RegionID', 'SizeRank', 'ZipCode', 'RegionType',
               'StateName', 'State', 'City', 'Metro', 'CountyName','Neighborhood',
               'HasUniversity')
date_cols <- setdiff(colnames(df), meta_cols)
```

```{r}
test_group_specificity <- function(data, response, effect, groups) {
  results <- list()
  
  for (g in groups) {
    # Base model: control for group + effect
    base_formula <- as.formula(
      paste(response, "~", g, "+", effect)
    )
    
    # Interaction model: allow effect to vary by group
    interaction_formula <- as.formula(
      paste(response, "~", g, "+", effect, "*", g)
    )
    
    model_base <- lm(base_formula, data = data)
    model_interaction <- lm(interaction_formula, data = data)
    
    cmp <- anova(model_base, model_interaction)
    
    results[[g]] <- data.frame(
      group = g,
      df_added = cmp$Df[2],
      rss_drop = cmp$RSS[1] - cmp$RSS[2],
      f_value = cmp$F[2],
      p_value = cmp$`Pr(>F)`[2]
    )
  }
  
  do.call(rbind, results)
}
```


## Converting data to long form
```{r}
long_df <- df %>%
  pivot_longer(cols = all_of(date_cols), 
               names_to = 'date', values_to = 'zorf') %>%
  mutate(date = as.Date(gsub('\\.', '-', gsub('^X', '', date)))) %>%
  arrange(ZipCode, date) %>%
  filter(!is.na(zorf), !is.na(date)) %>%
  mutate(
    HasUniversity = as.integer(HasUniversity),
    Neighborhood  = as.factor(Neighborhood)
  )
```

## EDA/ Feature Engineering

### Do weather patterns (seasons) impact ZORF?
```{r}
long_df <- long_df %>%
  mutate(
    month_num  = as.numeric(format(date, '%m')),
    sin_season = sin(2 * pi * month_num / 12),
    cos_season = cos(2 * pi * month_num / 12)
  )
```

```{r}
#anova checking the significance of including seasonality vs. without
lm_without <- lm(zorf ~ time_index, data = long_df)

lm_with <- lm(zorf ~ time_index + sin_season + cos_season, data = long_df)

anova(lm_without, lm_with)
```
```{r}
cor(long_df$sin_season, long_df$zorf, use = "complete.obs")
cor(long_df$cos_season, long_df$zorf, use = "complete.obs")
```
**Data Decision:** Since the joint F-test is significant but individual correlations are near zero, the seasonal terms matter only as a combined pattern, not as separate effects.

Is it different by group?
```{r}
groups <- c("HasUniversity", "Neighborhood", "ZipCode")

test_group_specificity(
  data = long_df,
  response = "zorf",
  effect = "sin_season + cos_season",
  groups = groups
)
```
The smooth sinusoidal seasonality (sin/cos) is globally consistent across locations, but it does not capture the true leasing cycle pattern.


### Does the academic calendar impact ZORF? Is it global or zipcode level?
To understand how best to model the “September 1” lease cycle in Boston, we first tested whether rent varies systematically by month without imposing any specific seasonal shape. We fit a linear regression model where zorf is explained by month categories.
```{r}
long_df$month_factor <- factor(long_df$month_num)

lm(zorf ~ month_factor, data = long_df)
```
The intercept represents average rent in January. Each month_factor coefficient tells you how much higher or lower that month is compared to January.By month shows prices rise from spring to a summer peak (June–July) and then decline after September, confirming a pre-9/1 demand cycle.
Given this, to model the leasing cycle effect, we construct a piecewise function to mimic that cycle.

```{r}
create_lease_cycle <- function(months) {
  sapply(months, function(m) {
    if (is.na(m)) return(0)

    # Jan-Apr: low
    if (m <= 4) return(0)

    # May-Jul: ramp up to peak
    if (m >= 5 && m <= 7) return((m - 4) / 3)

    # Aug-Sep: decline from peak
    if (m >= 8 && m <= 9) return(1 - 0.35 * (m - 7))

    # Oct-Dec: low again
    return(0)
  })
}

long_df$lease_cycle <- create_lease_cycle(long_df$month_num)
```

```{r}
groups <- c("HasUniversity", "Neighborhood", "ZipCode")

test_group_specificity(
  data = long_df,
  response = "zorf",
  effect = "lease_cycle",
  groups = groups
)
```
There are no significant differences in how ZORF follows the lease-cycle pattern across locations, meaning the seasonal effect is global rather than group-specific.

### Are there significant differences in long-term trend over time? Are some zips appreciating faster than others over time?
```{r}
summary(lm(zorf ~ time_index, data = long_df))
```
Time is statistically signficant, which makes sense given inflation on rent prices over time. 

```{r}
groups <- c("HasUniversity", "Neighborhood", "ZipCode")

test_group_specificity(
  data = long_df,
  response = "zorf",
  effect = "time_index",
  groups = groups
)
```
The rate of appreciation over time is significantly different by zip code and by neighborhood, but is consistent between university and non-university areas (p = 0.65).

```{r}
zip_trends <- long_df %>%
  group_by(ZipCode) %>%
  do(tidy(lm(zorf ~ time_index, data = .))) %>%
  filter(term == "time_index") %>%
  select(ZipCode, estimate, std.error, p.value) %>%
  arrange(desc(estimate))
```

```{r}
ggplot(long_df, aes(x = time_index, y = zorf, 
                    group = ZipCode, color = Neighborhood)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  theme_minimal() +
  labs(title = "Zip-level trends over time (colored by Neighborhood)", 
       x = "Time Index", y = "ZORF")
```
Looking at trends in ZORF over time though, most of the lines follow a similar slope, just starting intercepts are different. There are two outliers.


So, next testing whether neighborhood labels explain a statistically significant portion of the variation in zip-level time slopes.
```{r}
# get zip-level slopes
zip_slopes <- long_df %>%
  group_by(Neighborhood, ZipCode) %>%
  do(tidy(lm(zorf ~ time_index, data = .))) %>%
  filter(term == "time_index") %>%
  select(Neighborhood, ZipCode, estimate)
# Null: zip slopes are just randomly distributed around a global mean
m_null <- lm(estimate ~ 1, data = zip_slopes)

# Alternative: zip slopes vary systematically by neighborhood
m_neighborhood <- lm(estimate ~ Neighborhood, data = zip_slopes)

anova(m_null, m_neighborhood)
```
Yes, knowing which neighborhood a zip code belongs to tells you something meaningful about how fast that zip's rents are appreciating over time.

### Volatility
```{r}
calculate_volatility <- function(data, window = 6) {
  data %>%
    arrange(ZipCode, date) %>%
    group_by(ZipCode) %>%
    mutate(volatility = zoo::rollapply(
      zorf, 
      width   = window, 
      FUN     = sd, 
      fill    = NA, 
      align   = "right"   # only looks backwards
    )) %>%
    ungroup()
}

long_df <- calculate_volatility(long_df)
```

```{r}
# is it significant at all?
summary(lm(zorf ~ volatility, data = long_df))
```


```{r}
# is it global or group specific?
test_group_specificity(long_df %>% filter(!is.na(volatility)),
                       response = "zorf",
                       effect   = "volatility",
                       groups   = c("HasUniversity", "Neighborhood", "ZipCode"))
```



```{r}
zip_volatility_slopes <- long_df %>%
  filter(!is.na(volatility)) %>%
  group_by(Neighborhood, ZipCode) %>%
  do(tidy(lm(zorf ~ volatility, data = .))) %>%
  filter(term == "volatility") %>%
  select(Neighborhood, ZipCode, estimate) %>%
  left_join(distinct(long_df, ZipCode, HasUniversity), by = "ZipCode")

# does neighborhood explain zip-level slope variance?
anova(
  lm(estimate ~ 1,            data = zip_volatility_slopes),
  lm(estimate ~ Neighborhood, data = zip_volatility_slopes)
)
```
Neighborhood explains a statistically significant portion of the variation in how strongly volatility affects zip-level rents (F=2.83, p=0.035), justifying volatility as a neighborhood-level random slope rather than a global fixed effect.


**Does neighborhood explain additional slope variance beyond what HasUniversity already captures?** 
```{r}
anova(
  lm(estimate ~ HasUniversity,                data = zip_volatility_slopes),
  lm(estimate ~ HasUniversity + Neighborhood, data = zip_volatility_slopes)
)
```
So neighborhood does not significantly explain additional variance beyond what HasUniversity already captures (p=0.076)




## Train - Test Split
```{r}
cutoff_date <- quantile(long_df$date, 0.8)

train_df <- long_df %>% filter(date <= cutoff_date)
test_df  <- long_df %>% filter(date >  cutoff_date)

cat('nrow train_df:', nrow(train_df), '\n')
cat('nrow test_df:',  nrow(test_df),  '\n')
cat('train date range:', as.character(min(train_df$date)), 
    'to', as.character(max(train_df$date)), '\n')
cat('test date range:',  as.character(min(test_df$date)),  
    'to', as.character(max(test_df$date)),  '\n')
```



## Designing The Bayesian model
- Sine + Cosine interaction level is a global variable.
- lease_cycle is a global variable
- Knowing which neighborhood a zip code belongs to tells you something meaningful about how fast that zip's rents are appreciating over time.
- Has_university is a good indicator of volatility

```{r}
colnames(long_df)
```

```{r}
# initial priors
default_prior(zorf ~ time_index + sin_season + cos_season +
                HasUniversity + (1 | ZipCode) + (1 | Neighborhood),
              data   = train_df,
              family = Gamma(link = 'log'))
```

```{r}
my_prior <- c(
  prior(normal(0, 1),   class = 'b', coef = 'time_index'),
  prior(normal(0, 1),   class = 'b', coef = 'sin_season'),
  prior(normal(0, 1),   class = 'b', coef = 'cos_season'),
  prior(normal(0, 1),   class = 'b', coef = 'HasUniversity'),
  prior(normal(7.5, 0.5), class = 'Intercept'),
  prior(gamma(2, 0.2), class = 'shape'),
  prior(exponential(1), class = 'sd', group = 'ZipCode')
)
```



### fitting the model
```{r}
fit <- brm(
  zorf ~ time_index + sin_season + cos_season + 
         HasUniversity + (1 | ZipCode),
  data    = train_df,
  family  = Gamma(link = 'log'),
  prior   = my_prior,
  chains  = 4,
  iter    = 4000,
  warmup  = 1000,
  cores   = 4,
  seed    = 42,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```
```{r}
fit_lognormal <- brm(
  zorf ~ time_index + sin_season + cos_season +
         HasUniversity + (1 | ZipCode),
  data    = train_df,
  family  = lognormal(),
  prior   = c(
    prior(normal(0, 1), class = "b"),
    prior(normal(7.5, 0.5), class = "Intercept"),
    prior(exponential(1), class = "sd", group = "ZipCode"),
    prior(exponential(1), class = "sigma")
  ),
  chains  = 4,
  iter    = 4000,
  warmup  = 1000,
  cores   = 4,
  seed    = 42,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```

```{r}
summary(fit_lognormal)
```

```{r}
bayes_R2(fit_lognormal)
```

```{r}
pred_test2 <- posterior_predict(
  fit_lognormal,
  newdata          = test_df,
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian"
)  
preds_test_mean <- colMeans(pred_test2)
cor(preds_test_mean, test_df$zorf)
rmse_test <- sqrt(mean((preds_test_mean - test_df$zorf)^2))
rmse_test
# Check 90% CI coverage
lower <- apply(pred_test2, 2, quantile, 0.05)
upper <- apply(pred_test2, 2, quantile, 0.95)
coverage <- mean(test_df$zorf >= lower & test_df$zorf <= upper)
cat("90% CI coverage:", coverage, "\n")
```



## Evaluation
```{r}
summary(fit)
```



```{r}
plot(fit)
```


```{r}
bayes_R2(fit)
```


```{r}
post_preds_train <- posterior_predict(fit)
preds_train_mean <- colMeans(post_preds_train)
```


```{r}
plot(preds_train_mean, train_df$zorf,
     main = "Posterior Pred Means vs. Actual zorf (Train)",
     xlab = "Predicted zorf", ylab = "Actual zorf")
abline(0, 1, col = 'red', lwd = 2)   # perfect prediction line
```

```{r}
cor(preds_train_mean, train_df$zorf)
```


```{r}
post_preds_test <- posterior_predict(
  fit,
  newdata          = test_df,
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian"   # draws new group effects from the estimated
)                                  # population distribution (Σ_ψ) rather than
                                   # just using the fixed-effect mean
```

```{r}
preds_test_mean <- colMeans(post_preds_test)
cor(preds_test_mean, test_df$zorf)
```

```{r}
plot(preds_test_mean, test_df$zorf,
     main = "Posterior Pred Means vs. Actual zorf (Test)",
     xlab = "Predicted zorf", ylab = "Actual zorf")
abline(0, 1, col = 'red', lwd = 2)
```


```{r}
rmse_test <- sqrt(mean((preds_test_mean - test_df$zorf)^2))
rmse_test
```

```{r}
preds_test_mean <- colMeans(post_preds_test)
cor(preds_test_mean, test_df$zorf)
rmse_test <- sqrt(mean((preds_test_mean - test_df$zorf)^2))
rmse_test
# Check 90% CI coverage
lower <- apply(post_preds_test, 2, quantile, 0.05)
upper <- apply(post_preds_test, 2, quantile, 0.95)
coverage <- mean(test_df$zorf >= lower & test_df$zorf <= upper)
cat("90% CI coverage:", coverage, "\n")
```

### Testing how adding other enginered features impacts results 
```{r}
my_prior <- c(
  prior(normal(0, 1),   class = 'b', coef = 'time_index'),
  prior(normal(0, 1),   class = 'b', coef = 'sin_season'),
  prior(normal(0, 1),   class = 'b', coef = 'cos_season'),
  prior(normal(0, 1),   class = 'b', coef = 'HasUniversity'),
  prior(normal(7.5, 0.5), class = 'Intercept'),
  prior(gamma(2, 0.1),     class = 'shape'),
  prior(exponential(1), class = 'sd', group = 'ZipCode'),
  prior(normal(0, 1), class = 'b', coef = 'sept_intensity')
)
```

```{r}
fit2 <- brm(
  zorf ~ time_index + sin_season + cos_season + 
       HasUniversity + sept_intensity + (1 | ZipCode),
  data    = train_df,
  family  = Gamma(link = 'log'),
  prior   = my_prior,
  chains  = 4,
  iter    = 4000,
  warmup  = 1000,
  cores   = 4,
  seed    = 42,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```

```{r}
summary(fit2)
```


```{r}
plot(fit2)
```


```{r}
bayes_R2(fit2)
```


```{r}
pred_test2 <- posterior_predict(
  fit,
  newdata          = test_df,
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian"   # draws new group effects from the estimated
)    
```


```{r}
pred_test2 <- posterior_predict(
  fit2,
  newdata          = test_df,
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian"   # draws new group effects from the estimated
)  
preds_test_mean <- colMeans(pred_test2)
cor(preds_test_mean, test_df$zorf)
rmse_test <- sqrt(mean((preds_test_mean - test_df$zorf)^2))
rmse_test
# Check 90% CI coverage
lower <- apply(pred_test2, 2, quantile, 0.05)
upper <- apply(pred_test2, 2, quantile, 0.95)
coverage <- mean(test_df$zorf >= lower & test_df$zorf <= upper)
cat("90% CI coverage:", coverage, "\n")
```




# NEXT TEST:
```{r}
my_prior2 <- c(
  prior(normal(0, 1), class = "b", coef = "time_index"),
  prior(normal(0, 1), class = "b", coef = "sin_season"),
  prior(normal(0, 1), class = "b", coef = "cos_season"),
  prior(normal(0, 1), class = "b", coef = "HasUniversity"),
  prior(normal(0, 1), class = "b", coef = "sept_intensity"),
  prior(normal(7.5, 0.5), class = "Intercept"),

  # mean-model random effect SD
  prior(exponential(1), class = "sd", group = "ZipCode"),

  # shape-model intercept
  prior(normal(3, 1), class = "Intercept", dpar = "shape"),

  # shape-model random effect SD
  prior(exponential(1), class = "sd", group = "ZipCode", dpar = "shape")
)
```


```{r}
fit2 <- brm(
  bf(
    zorf ~ time_index + sin_season + cos_season +
      HasUniversity + sept_intensity + (1 | ZipCode),
    shape ~ 1 + (1 | ZipCode)
  ),
  data    = train_df,
  family  = Gamma(link = "log"),
  prior   = my_prior2,
  chains  = 4,
  iter    = 4000,
  warmup  = 1000,
  cores   = 4,
  seed    = 42,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```


```{r}
my_prior2 <- c(
  prior(normal(0, 1), class = "b", coef = "time_index"),
  prior(normal(0, 1), class = "b", coef = "sin_season"),
  prior(normal(0, 1), class = "b", coef = "cos_season"),
  prior(normal(0, 1), class = "b", coef = "HasUniversity"),
  prior(normal(0, 1), class = "b", coef = "sept_intensity"),
  prior(normal(7.5, 0.5), class = "Intercept"),

  prior(exponential(1), class = "sd", group = "ZipCode"),

  prior(normal(3, 1), class = "Intercept", dpar = "shape"),
  prior(exponential(1), class = "sd", group = "ZipCode", dpar = "shape")
)

fit2 <- brm(
  bf(
    zorf ~ time_index + sin_season + cos_season +
      HasUniversity + sept_intensity + (1 | ZipCode),
    shape ~ 1 + (1 | ZipCode)
  ),
  data    = train_df,
  family  = Gamma(link = "log"),
  prior   = my_prior2,
  chains  = 4,
  iter    = 4000,
  warmup  = 1000,
  cores   = 4,
  seed    = 42,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```


```{r}
summary(fit2)
plot(fit2)
bayes_R2(fit2)

pred_test2 <- posterior_predict(
  fit2,
  newdata            = test_df,
  allow_new_levels   = TRUE,
  sample_new_levels  = "gaussian"
)

preds_test_mean <- colMeans(pred_test2)
cor(preds_test_mean, test_df$zorf)

rmse_test <- sqrt(mean((preds_test_mean - test_df$zorf)^2))
rmse_test

lower <- apply(pred_test2, 2, quantile, 0.05)
upper <- apply(pred_test2, 2, quantile, 0.95)
coverage <- mean(test_df$zorf >= lower & test_df$zorf <= upper)

cat("90% CI coverage:", coverage, "\n")
```

```{r}
fitted_vals <- preds_test_mean
residuals <- test_df$zorf - fitted_vals

plot(fitted_vals, residuals,
     pch = 16, col = rgb(0,0,0,0.3),
     xlab = "Fitted", ylab = "Residuals")
abline(h = 0, col = "red")
```

```{r}
plot(fitted_vals, abs(residuals),
     pch = 16, col = rgb(0,0,0,0.3),
     xlab = "Fitted", ylab = "|Residual|")
```

```{r}
plot(test_df$time_index, residuals,
     pch = 16, col = rgb(0,0,0,0.3),
     xlab = "Time", ylab = "Residuals")
abline(h = 0, col = "red")
```


```{r}
covered <- (test_df$zorf >= lower & test_df$zorf <= upper)

coverage_by_zip <- aggregate(covered ~ ZipCode, data = test_df, mean)
plot(coverage_by_zip$covered,
     main = "Coverage by ZipCode", ylab = "Coverage")
```


```{r}
colnames(long_df)
```

```{r}
coverage_by_neighborhood <- aggregate(covered ~ Neighborhood, data = test_df, mean)
```


```{r}
ggplot(long_df, aes(x = time_index, y = zorf, color = ZipCode)) +
  geom_line(alpha = 0.3)
```

```{r}
residuals <- test_df$zorf - preds_test_mean
var_by_neighborhood <- aggregate(residuals ~ Neighborhood, data = test_df, function(x) var(x))
```



```{r}
long_df %>%
  group_by(ZipCode) %>%
  summarise(slope = coef(lm(zorf ~ time_index))[2]) %>%
  summary()
```

```{r}
sample_neighborhoods <- long_df %>%
  distinct(Neighborhood) %>%
  slice(1:6) %>%
  pull()

ggplot(filter(long_df, Neighborhood %in% sample_neighborhoods),
       aes(x = time_index, y = zorf, color = ZipCode)) +
  geom_line(size = 1) +
  facet_wrap(~ Neighborhood, scales = "free_y") +
  theme_minimal()
```


```{r}
sample_neighborhoods <- long_df %>%
  distinct(Neighborhood) %>%
  slice(7:16) %>%
  pull()

ggplot(filter(long_df, Neighborhood %in% sample_neighborhoods),
       aes(x = time_index, y = zorf, color = ZipCode)) +
  geom_line(size = 1) +
  facet_wrap(~ Neighborhood, scales = "free_y") +
  theme_minimal()
```

```{r}

```




```{r}
ggplot(long_df, aes(x = time_index, y = zorf, color = factor(HasUniversity))) +
  geom_line(alpha = 0.4) +
    facet_wrap(~ Neighborhood, scales = "free_y") +
  theme_minimal()
```

```{r}
neighs <- unique(long_df$Neighborhood)
# split into two halves
first_half <- neighs[1:floor(length(neighs)/2)]
second_half <- neighs[(floor(length(neighs)/2)+1):length(neighs)]

# ---- Plot 1 ----
ggplot(filter(long_df, Neighborhood %in% first_half),
       aes(x = time_index, y = zorf, color = factor(HasUniversity))) +
  geom_line(alpha = 0.4) +
  facet_wrap(~ Neighborhood, scales = "free_y") +
  theme_minimal()

# ---- Plot 2 ----
ggplot(filter(long_df, Neighborhood %in% second_half),
       aes(x = time_index, y = zorf, color = factor(HasUniversity))) +
  geom_line(alpha = 0.4) +
  facet_wrap(~ Neighborhood, scales = "free_y") +
  theme_minimal()
```


```{r}
residuals <- long_df$zorf - fitted_vals

ggplot(long_df, aes(x = Neighborhood, y = residuals)) +
  geom_boxplot()
```


```{r}

long_df %>%
  group_by(Neighborhood, ZipCode) %>%
  summarise(slope = coef(lm(zorf ~ time_index))[2]) %>%
  ggplot(aes(x = Neighborhood, y = slope)) +
  geom_boxplot()
```


```{r}
library(lmtest)
bptest(lm(zorf ~ time_index, data = long_df))
```

```{r}
anova(lm(zorf ~ time_index + HasUniversity, data = long_df),
      lm(zorf ~ time_index * HasUniversity, data = long_df))
```

```{r}
anova(lm(zorf ~ time_index, data = long_df),
      lm(zorf ~ time_index * Neighborhood, data = long_df))
```

```{r}
anova(lm(zorf ~ time_index, data = long_df),
      lm(zorf ~ time_index * ZipCode, data = long_df))
```



```{r}
anova(
  lm(zorf ~ sept_intensity + HasUniversity, data = long_df),
  lm(zorf ~ sept_intensity * HasUniversity, data = long_df)
)
```



```{r}
library(dplyr)
library(purrr)
library(tibble)

# ---------------------------
# 1) Choose variables to test
# ---------------------------
vars_to_test <- c(
  "time_index",
  "sin_season",
  "cos_season",
  "season_flag",
  "sept_intensity",
  "rolling_volatility",
  "HasUniversity",
  "Neighborhood",
  "ZipCode"
)

# Keep only variables that actually exist in long_df
vars_to_test <- vars_to_test[vars_to_test %in% names(long_df)]

# Optional: convert grouping vars to factors if they are not already
long_df <- long_df %>%
  mutate(
    HasUniversity = as.factor(HasUniversity),
    Neighborhood  = as.factor(Neighborhood),
    ZipCode       = as.factor(ZipCode)
  )

# -----------------------------------
# 2) Fit additive "global" base model
# -----------------------------------
base_formula <- as.formula(
  paste("zorf ~", paste(vars_to_test, collapse = " + "))
)

base_model <- lm(base_formula, data = long_df)

# ------------------------------------------------
# 3) Generate all pairwise interactions to compare
# ------------------------------------------------
interaction_pairs <- combn(vars_to_test, 2, simplify = FALSE)

# -------------------------------------------------------
# 4) For each pair: fit additive + one interaction model
# -------------------------------------------------------
interaction_results <- map_dfr(interaction_pairs, function(pair) {
  v1 <- pair[1]
  v2 <- pair[2]

  interaction_term <- paste0(v1, ":", v2)

  interaction_formula <- as.formula(
    paste(
      "zorf ~",
      paste(vars_to_test, collapse = " + "),
      "+",
      interaction_term
    )
  )

  interaction_model <- lm(interaction_formula, data = long_df)

  cmp <- anova(base_model, interaction_model)

  tibble(
    var1 = v1,
    var2 = v2,
    interaction = interaction_term,
    df_added = cmp$Df[2],
    rss_base = cmp$RSS[1],
    rss_interaction = cmp$RSS[2],
    rss_drop = cmp$RSS[1] - cmp$RSS[2],
    f_value = cmp$F[2],
    p_value = cmp$`Pr(>F)`[2]
  )
})

# -----------------------------------------
# 5) Adjust p-values for multiple testing
# -----------------------------------------
interaction_results <- interaction_results %>%
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    p_fdr = p.adjust(p_value, method = "fdr")
  ) %>%
  arrange(p_fdr, p_value)

# ---------------------------
# 6) View the ranked results
# ---------------------------
print(interaction_results, n = Inf)

# Optional: only interactions that survive FDR
significant_interactions <- interaction_results %>%
  filter(!is.na(p_fdr), p_fdr < 0.05)

print(significant_interactions, n = Inf)
```



```{r}
colnames(long_df)
```


```{r}
library(tidyverse)
library(broom)

results <- long_df %>%
  group_by(Neighborhood) %>%
  # Filter for neighborhoods with at least 2 unique ZipCodes
  filter(n_distinct(ZipCode) > 1) %>%
  nest() %>%
  mutate(
    # Fit the two models
    mod_parallel = map(data, ~lm(zorf ~ time_index + ZipCode, data = .x)),
    mod_interaction = map(data, ~lm(zorf ~ time_index * ZipCode, data = .x)),
    
    # Compare them using ANOVA
    comparison = map2(mod_parallel, mod_interaction, ~anova(.x, .y)),
    
    # Extract the p-value
    p_value = map_dbl(comparison, ~ .x$`Pr(>F)`[2])
  ) %>%
  select(Neighborhood, p_value) %>%
  mutate(significant_difference = p_value < 0.05)
```


```{r}
table(results$significant_difference)
```


```{r}
# Extract the slopes for the neighborhoods with significant differences
divergent_names <- results %>% filter(significant_difference == TRUE) %>% pull(Neighborhood)

long_df %>%
  filter(Neighborhood %in% divergent_names) %>%
  group_by(Neighborhood, ZipCode) %>%
  do(tidy(lm(zorf ~ time_index, data = .))) %>%
  filter(term == "time_index") %>%
  select(Neighborhood, ZipCode, estimate) # 'estimate' is the growth rate
```

