---
title: "bayesion model"
author: "Krithika Natarajan"
date: "2026-04-09"
output: pdf_document
---

```{r setup, include=FALSE}
install.packages(c('tidyr', 'dplyr'))
library(tidyr)
library(dplyr)
```

## Reading in the Data
```{r}
df <- read.csv('boston_cleaned_data.csv', colClasses = c(ZipCode = 'character'))
 
meta_cols <- c('RegionID', 'SizeRank', 'ZipCode', 'RegionType',
               'StateName', 'State', 'City', 'Metro', 'CountyName')
date_cols <- setdiff(colnames(df), meta_cols)
```

## Converting data to long form
```{r}
long_df <- df %>%
  pivot_longer(cols = all_of(date_cols), names_to = 'date', values_to = 'zorf') %>%
  mutate(date = as.Date(gsub('\\.', '-', gsub('^X', '', date)))) %>%
  arrange(ZipCode, date) %>%
  filter(!is.na(zorf), !is.na(date)) %>%  # filter first
  mutate(                                  # then add features
    month               = as.integer(format(date, '%m')),
    is_before_september = as.integer(month < 9)
  )
```

## EDA

### Test 1: How do different zipcode rental prices change over time?
```{r}
long_df %>%
  group_by(ZipCode) %>%
  summarise(mean_zorf = mean(zorf)) %>%
  arrange(desc(mean_zorf))
```

```{r}
long_df %>%
  ggplot(aes(x = date, y = zorf, color = ZipCode)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'ZORF over time by Zip Code')
```
Lines follow the same patterns over time (all dip during covid) and looking at seasonality, seasonal and market trends are shared across zip codes.
However the baseline in which they operate are drastically different.


```{r}
train_df %>%
  ggplot(aes(x = reorder(ZipCode, zorf, median), y = zorf)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  labs(title = 'ZORF distribution by Zip Code',
       x = 'Zip Code', y = 'ZORF')
```


```{r}
anova_result <- aov(zorf ~ ZipCode, data = long_df)
summary(anova_result)
```
One-way ANOVA tests whether the variation in ZORF between zip codes is large enough relative to the variation within zip codes to be considered statistically significant
With a result of (F=146.7, p<2e-16) means the zip code differences are overwhelmingly real and not due to chance, justifying the use of zip-level random effects in the model.


### Introducing Seasonality

```{r}
long_df <- long_df %>% arrange(date)
```

```{r}

# Recompute seasonality after sorting just to be safe
month_num <- as.numeric(format(long_df$date, "%m"))
long_df$sin_season <- sin(2 * pi * month_num / 12)
long_df$cos_season <- cos(2 * pi * month_num / 12)
```


## Building the model

```{r}
# Train/test split
split_idx <- floor(nrow(long_df) * 0.8)
train_df  <- long_df %>% slice(1:split_idx)
test_df   <- long_df %>% slice((split_idx + 1):nrow(long_df))
cat('nrow train_df:', nrow(train_df), '\n')
cat('nrow test_df:',  nrow(test_df),  '\n')
```


### Zip-level Random Effects
```{r}
zip_means <- train_df %>%
  group_by(ZipCode) %>%
  summarise(
    zip_mean = mean(zorf),
    n_obs    = n()  # keep track of how many observations
                    # each zip code contributed — useful for
                    # diagnosing which zip codes are sparse
  )

cat('Zip code observation counts:\n')
print(summary(zip_means$n_obs))
```


```{r}
overall_mean <- mean(train_df$zorf)

# Join zip means into training data
train_df <- train_df %>%
  left_join(zip_means %>% dplyr::select(ZipCode, zip_mean), 
            by = 'ZipCode')

# Join zip means into test data
# Any zip code in test that wasn't seen in training
# gets the overall training mean as its baseline.
# This is conservative — we're saying we have no
# neighborhood-specific information for this zip code
# so we assume it behaves like an average neighborhood.
test_df <- test_df %>%
  left_join(zip_means %>% dplyr::select(ZipCode, zip_mean), 
            by = 'ZipCode') %>%
  mutate(zip_mean = ifelse(is.na(zip_mean), overall_mean, zip_mean))

# Check how many test zip codes needed the fallback
n_fallback <- sum(is.na(test_df$zip_mean) | 
                  test_df$zip_mean == overall_mean)
cat('Test observations using fallback mean:', n_fallback, '\n')
```


### Training the Model
```{r}
# Build y and X for training
# Subtract zip-level mean from y.
# The model now learns seasonal patterns and 
# residual variance on top of the neighborhood baseline,
# rather than having to explain both at once.
# The intercept in X should be close to zero on average
# since the zip means have already absorbed most of
# the level differences across neighborhoods.
y <- train_df$zorf - train_df$zip_mean

X <- cbind(
  intercept  = 1,
  sin_season = train_df$sin_season,
  cos_season = train_df$cos_season
)

n <- nrow(X)
p <- ncol(X)

cat('X dimensions:', n, 'x', p, '\n')
cat('y mean after zip adjustment:', mean(y), '\n')
cat('y sd after zip adjustment:',  sd(y),   '\n')
```


```{r}
# Priors for w
# Zero prior mean for all three weights:
# - Intercept: should be near zero after zip adjustment
# - sin/cos: no seasonality is a reasonable null assumption,
#   the data will push these away from zero if seasonality exists
# Prior variance of 10 is relatively uninformative —
# we're not strongly penalizing large seasonal effects
prior_w     <- matrix(rep(0, p), ncol = 1)
prior_Sigma <- diag(p) * 10

# Priors for sigma2
# Alpha = 3 is a weakly informative shape
# Beta anchored to empirical variance of adjusted y
# so the prior is on the right scale for your data
prior_alpha <- 3
prior_beta  <- var(y) * 2
```

```{r}
# Storage and initialization
n_samples <- 10000
gibbs_samples_w      <- matrix(0, nrow = n_samples, ncol = p)
gibbs_samples_sigma2 <- matrix(0, nrow = n_samples, ncol = 1)

# Initialize at sensible starting values
w_tilde      <- matrix(0, nrow = p, ncol = 1)
sigma2_tilde <- var(y)


for (i in 1:n_samples) {
  # Conditional posterior for w given sigma2
  V       <- solve((1/sigma2_tilde) * t(X) %*% X + solve(prior_Sigma))
  M       <- V %*% ((1/sigma2_tilde) * t(X) %*% y + 
                    solve(prior_Sigma) %*% prior_w)
  L       <- chol(V)
  w_tilde <- M + t(L) %*% rnorm(p)

  # Conditional posterior for sigma2 given w
  A            <- prior_alpha + n / 2
  B            <- prior_beta + 0.5 * sum((y - X %*% w_tilde)^2)
  
  # Sample from inverse gamma using base R only
  sigma2_tilde <- 1 / rgamma(1, shape = A, rate = B)

  gibbs_samples_w[i, ]      <- w_tilde
  gibbs_samples_sigma2[i, ] <- sigma2_tilde
}

cat('Posterior weight means:\n')
print(colMeans(gibbs_samples_w))
cat('Posterior sigma2 mean:\n')
print(mean(gibbs_samples_sigma2))
```

```{r}
# Convergence diagnostics
# Always check these before trusting the posterior samples.
# The trace plots should look like white noise with no drift.
# The ACF should drop to near zero quickly.
# If either looks bad, you may need more samples or thinning.
par(mfrow = c(2, 3))
plot(gibbs_samples_w[,1],      type = 'l', main = 'Trace: Intercept')
plot(gibbs_samples_w[,2],      type = 'l', main = 'Trace: sin')
plot(gibbs_samples_w[,3],      type = 'l', main = 'Trace: cos')
plot(gibbs_samples_sigma2,     type = 'l', main = 'Trace: sigma2')
acf(gibbs_samples_w[,2],                   main = 'ACF: sin')
acf(gibbs_samples_sigma2,                  main = 'ACF: sigma2')
par(mfrow = c(1, 1))
```

### Evaluation
```{r}
# Posterior predictive for test set
# Build the test X matrix the same way as training
y_test <- test_df$zorf - test_df$zip_mean
X_test <- cbind(
  intercept  = 1,
  sin_season = test_df$sin_season,
  cos_season = test_df$cos_season
)

n_test          <- nrow(X_test)
post_pred_means <- numeric(n_test)
post_pred_lower <- numeric(n_test)
post_pred_upper <- numeric(n_test)

for (j in 1:n_test) {
  # For each test observation, generate a full
  # posterior predictive distribution by plugging
  # each posterior draw of w and sigma2 into the likelihood
  preds <- rnorm(
    n    = n_samples,
    mean = gibbs_samples_w %*% X_test[j, ],
    sd   = sqrt(gibbs_samples_sigma2)
  )
  post_pred_means[j] <- mean(preds)
  # 90% credible interval — the Bayesian analog of a
  # confidence interval, directly interpretable as
  # "we believe true rent falls in this range with 90% probability"
  post_pred_lower[j] <- quantile(preds, 0.05)
  post_pred_upper[j] <- quantile(preds, 0.95)
}

# Add zip mean back to get predictions on original rent scale
final_preds <- post_pred_means + test_df$zip_mean
final_lower <- post_pred_lower  + test_df$zip_mean
final_upper <- post_pred_upper  + test_df$zip_mean

# Evaluation
rmse <- sqrt(mean((final_preds - test_df$zorf)^2))
coverage <- mean(test_df$zorf >= final_lower & 
                 test_df$zorf <= final_upper)

cat('Test RMSE:', rmse, '\n')
cat('90% credible interval coverage:', coverage, '\n')
```

```{r}
# Choose your zip code of interest
target_zip <- "02115"  # example — Fenway/Northeastern area

# Check if this zip code was in training data
if (!target_zip %in% zip_means$ZipCode) {
  cat('Warning: zip code not seen in training, using overall mean\n')
  zip_baseline <- overall_mean
} else {
  zip_baseline <- zip_means$zip_mean[zip_means$ZipCode == target_zip]
}
cat('Zip baseline rent:', zip_baseline, '\n')
```

