---
title: "bayesion model"
author: "Krithika Natarajan"
date: "2026-04-09"
output: pdf_document
---

```{r setup, include=FALSE}
install.packages(c('tidyr', 'dplyr'))
library(tidyr)
library(dplyr)
```

## Reading in the Data
```{r}
df <- read.csv('boston_cleaned_data.csv', colClasses = c(ZipCode = 'character'))
 
meta_cols <- c('RegionID', 'SizeRank', 'ZipCode', 'RegionType',
               'StateName', 'State', 'City', 'Metro', 'CountyName')
date_cols <- setdiff(colnames(df), meta_cols)
```

## Converting data to long form
```{r}
long_df <- df %>%
  pivot_longer(cols = all_of(date_cols), names_to = 'date', values_to = 'zorf') %>%
  mutate(date = as.Date(gsub('\\.', '-', gsub('^X', '', date)))) %>%
  arrange(ZipCode, date) %>%
  filter(!is.na(zorf), !is.na(date)) %>%  # filter first
  mutate(                                  # then add features
    month               = as.integer(format(date, '%m')),
    is_before_september = as.integer(month < 9)
  )
```

## EDA

### Test 1: How do different zipcode rental prices change over time?
```{r}
long_df %>%
  group_by(ZipCode) %>%
  summarise(mean_zorf = mean(zorf)) %>%
  arrange(desc(mean_zorf))
```

```{r}
long_df %>%
  ggplot(aes(x = date, y = zorf, color = ZipCode)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'ZORF over time by Zip Code')
```
Lines follow the same patterns over time (all dip during covid) and looking at seasonality, seasonal and market trends are shared across zip codes.
However the baseline in which they operate are drastically different.


```{r}
train_df %>%
  ggplot(aes(x = reorder(ZipCode, zorf, median), y = zorf)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  labs(title = 'ZORF distribution by Zip Code',
       x = 'Zip Code', y = 'ZORF')
```


```{r}
anova_result <- aov(zorf ~ ZipCode, data = long_df)
summary(anova_result)
```
One-way ANOVA tests whether the variation in ZORF between zip codes is large enough relative to the variation within zip codes to be considered statistically significant
With a result of (F=146.7, p<2e-16) means the zip code differences are overwhelmingly real and not due to chance, justifying the use of zip-level random effects in the model.


## Building the model

```{r}
split_idx <- floor(nrow(long_df) * 0.8)
train_df <- long_df %>% slice(1:split_idx)
test_df  <- long_df %>% slice((split_idx + 1):nrow(long_df))

cat('nrow train_df:', nrow(train_df), '\n')
cat('nrow test_df:', nrow(test_df), '\n')
```


### Zip-level Random Effects
```{r}
grand_mean   <- mean(long_df$zorf)
sd_threshold <- sd(long_df$zorf)

zip_means <- long_df %>%
  group_by(ZipCode) %>%
  summarise(zip_mean = mean(zorf)) %>%
  mutate(
    deviation = zip_mean - grand_mean,
    tier      = case_when(
      deviation >  sd_threshold ~ 'high',
      deviation < -sd_threshold ~ 'low',
      TRUE                      ~ 'mid'
    )
  )
```
Zip codes were grouped into three tiers, high, mid, and low, based on whether their mean ZORF deviated from the grand mean by more than one standard deviation. This threshold was chosen because it reflects a statistically meaningful departure from typical Boston rent levels, and the resulting tiers align naturally with known neighborhood-specific market dynamics.

Adding those features back to our dataset
```{r}
# Step 1: join tier into train and test via ZipCode
train_df <- train_df %>%
  left_join(zip_means %>% dplyr::select(ZipCode, tier), by = 'ZipCode')
test_df <- test_df %>%
  left_join(zip_means %>% dplyr::select(ZipCode, tier), by = 'ZipCode')

# Compute tier_means from long_df (not train_df)
tier_means <- long_df %>%
  left_join(zip_means %>% dplyr::select(ZipCode, tier), by = 'ZipCode') %>%
  group_by(tier) %>%
  summarise(tier_mean = mean(zorf))

# Step 2: join tier_mean via tier
train_df <- train_df %>%
  left_join(tier_means, by = 'tier')
test_df <- test_df %>%
  left_join(tier_means, by = 'tier')
```

### Training the Model
```{r}
### Training the Model
FEATURES  <- c('tier_mean')
# FEATURES <- c(FEATURES, 'is_before_september', 'zordi', 'avg_temp')

n_samples <- 10000

# Center y by tier mean — model learns deviations from tier baseline
y <- train_df$zorf - train_df$tier_mean
X <- cbind(1, as.matrix(train_df %>% dplyr::select(all_of(FEATURES))))

n <- nrow(X)
p <- ncol(X)

w_hat        <- solve(t(X) %*% X) %*% t(X) %*% y
df_resid     <- n - p
scale_resid  <- (t(y - X %*% w_hat) %*% (y - X %*% w_hat) / df_resid)[1, 1]
chi_draws    <- rchisq(n_samples, df = df_resid)
sigma2_tilde <- (df_resid * scale_resid) / chi_draws

XtX_inv <- solve(t(X) %*% X)
w_tilde  <- matrix(0, nrow = n_samples, ncol = p)
for (i in 1:n_samples) {
  L            <- chol(XtX_inv * sigma2_tilde[i])
  w_tilde[i, ] <- w_hat + t(L) %*% rnorm(p)
}

cat('w_hat:\n');                  print(w_hat)
cat('Posterior weight means:\n'); print(colMeans(w_tilde))
```

### Evaluation
```{r}
### Evaluate
X_test <- cbind(1, as.matrix(test_df %>% dplyr::select(all_of(FEATURES))))
y_test <- test_df$zorf - test_df$tier_mean   # center test y the same way
y_pred <- X_test %*% colMeans(w_tilde)

# Add tier mean back to get predictions in original dollar scale
y_pred_dollars <- y_pred + test_df$tier_mean
y_test_dollars <- test_df$zorf

rmse <- sqrt(mean((y_pred_dollars - y_test_dollars)^2))
corr <- cor(y_pred_dollars, y_test_dollars)
cat(sprintf('Test RMSE: %.4f\n', rmse))
cat(sprintf('Test Correlation: %.4f\n', corr))
```


### Running predictions
```{r}
predict_zip <- function(zip_code) {
  zip_info  <- zip_means %>% filter(ZipCode == as.character(zip_code))

  if (nrow(zip_info) == 0) {
    cat(sprintf('Zip code %s not found.\n', zip_code))
    return(invisible(NULL))
  }

  tier_info <- tier_means %>% filter(tier == zip_info$tier)
  x_new     <- c(1, tier_info$tier_mean)

  preds <- sapply(1:n_samples, function(i) {
    rnorm(1, mean = t(w_tilde[i, ]) %*% x_new, sd = sqrt(sigma2_tilde[i]))
  })

  # Add tier mean back to convert from deviation to dollar scale
  preds_dollars <- preds + tier_info$tier_mean

  pred_mean <- mean(preds_dollars)
  ci        <- quantile(preds_dollars, c(0.025, 0.975))

  cat(sprintf('\nZip Code: %s\n', zip_code))
  cat(sprintf('Tier: %s\n', zip_info$tier))
  cat(sprintf('Predicted ZORF: $%.2f\n', pred_mean))
  cat(sprintf('95%% CI: [$%.2f, $%.2f]\n', ci[1], ci[2]))
}
```


```{r}
predict_zip('02115')
predict_zip('02148')
predict_zip('02120')
```

