---
title: "bayesion model"
author: "Krithika Natarajan"
date: "2026-04-09"
output: pdf_document
---

```{r setup, include=FALSE}
install.packages(c('tidyr', 'dplyr', 'brms', 'ggplot2'))
library(tidyr)
library(dplyr)
library(ggplot2)
library(brms)
```



## Reading in the Data

```{r}
df <- read.csv('boston_cleaned_data.csv', colClasses = c(ZipCode = 'character'))
 
meta_cols <- c('RegionID', 'SizeRank', 'ZipCode', 'RegionType',
               'StateName', 'State', 'City', 'Metro', 'CountyName','Neighborhood',
               'HasUniversity')
date_cols <- setdiff(colnames(df), meta_cols)
```

## Converting data to long form
```{r}
long_df <- df %>%
  pivot_longer(cols = all_of(date_cols), 
               names_to = 'date', values_to = 'zorf') %>%
  mutate(date = as.Date(gsub('\\.', '-', gsub('^X', '', date)))) %>%
  arrange(ZipCode, date) %>%
  filter(!is.na(zorf), !is.na(date)) %>%
  mutate(
    HasUniversity = as.integer(HasUniversity),
    Neighborhood  = as.factor(Neighborhood)
  )
```

## EDA

### Test 1: How do different zipcode rental prices change over time?
```{r}
long_df %>%
  group_by(ZipCode) %>%
  summarise(mean_zorf = mean(zorf)) %>%
  arrange(desc(mean_zorf))
```

```{r}
long_df %>%
  ggplot(aes(x = date, y = zorf, color = ZipCode)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'ZORF over time by Zip Code')
```
Lines follow the same patterns over time (all dip during covid) and looking at seasonality, seasonal and market trends are shared across zip codes.
However the baseline in which they operate are drastically different.


```{r}
long_df %>%
  ggplot(aes(x = reorder(ZipCode, zorf, median), y = zorf)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  labs(title = 'ZORF distribution by Zip Code',
       x = 'Zip Code', y = 'ZORF')
```


```{r}
anova_result <- aov(zorf ~ ZipCode, data = long_df)
summary(anova_result)
```
One-way ANOVA tests whether the variation in ZORF between zip codes is large enough relative to the variation within zip codes to be considered statistically significant
With a result of (F=146.7, p<2e-16) means the zip code differences are overwhelmingly real and not due to chance, justifying the use of zip-level random effects in the model.


### Introducing Seasonality

```{r}
long_df <- long_df %>%
  mutate(
    time_index = as.numeric(date - min(date)) / 365,
    month_num  = as.numeric(format(date, '%m')),
    sin_season = sin(2 * pi * month_num / 12),
    cos_season = cos(2 * pi * month_num / 12)
  )
```

## Building the model

```{r}
# Train/test split
split_idx <- floor(nrow(long_df) * 0.8)
train_df  <- long_df %>% slice(1:split_idx)
test_df   <- long_df %>% slice((split_idx + 1):nrow(long_df))
cat('nrow train_df:', nrow(train_df), '\n')
cat('nrow test_df:',  nrow(test_df),  '\n')
```

```{r}
# Do higher-rent zips have higher variance?
train_df %>%
  group_by(ZipCode) %>%
  summarise(mean_zorf = mean(zorf), sd_zorf = sd(zorf)) %>%
  ggplot(aes(x = mean_zorf, y = sd_zorf)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE) +
  theme_minimal() +
  labs(title = 'Mean vs SD of ZORF by Zip Code',
       x = 'Mean ZORF', y = 'SD ZORF')
```


```{r}
# initial priors
default_prior(zorf ~ time_index + sin_season + cos_season +
                HasUniversity + (1 | ZipCode) + (1 | Neighborhood),
              data   = train_df,
              family = Gamma(link = 'log'))
```

```{r}
my_prior <- c(
  prior(normal(0, 1),    class = 'b', coef = 'time_index'),
  prior(normal(0, 1),    class = 'b', coef = 'sin_season'),
  prior(normal(0, 1),    class = 'b', coef = 'cos_season'),
  prior(normal(0, 1),    class = 'b', coef = 'HasUniversity'),

  prior(normal(5, 1),    class = 'Intercept'),

  prior(exponential(1),  class = 'sd', group = 'Neighborhood'),
  prior(exponential(1),  class = 'sd', group = 'Neighborhood:ZipCode')
)
```

### fitting the model

```{r}
fit <- brm(
  zorf ~ time_index + sin_season + cos_season +
         HasUniversity + (1 | Neighborhood/ZipCode),
  data    = train_df,
  family  = Gamma(link = 'log'),
  prior   = my_prior,
  chains  = 4,
  iter    = 2500,
  warmup  = 500,
  cores   = 4,
  seed    = 42,
  control = list(adapt_delta = 0.95)
)
```

## Evaluation
```{r}
summary(fit)
```



```{r}
plot(fit)
```


```{r}
bayes_R2(fit)
```


```{r}
post_preds_train <- posterior_predict(fit)
preds_train_mean <- colMeans(post_preds_train)
```


```{r}
plot(preds_train_mean, train_df$zorf,
     main = "Posterior Pred Means vs. Actual zorf (Train)",
     xlab = "Predicted zorf", ylab = "Actual zorf")
abline(0, 1, col = 'red', lwd = 2)   # perfect prediction line
```

```{r}
cor(preds_train_mean, train_df$zorf)
```


```{r}
post_preds_test <- posterior_predict(
  fit,
  newdata          = test_df,
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian"   # draws new group effects from the estimated
)                                  # population distribution (Σ_ψ) rather than
                                   # just using the fixed-effect mean
```

```{r}
preds_test_mean <- colMeans(post_preds_test)
cor(preds_test_mean, test_df$zorf)
```

```{r}
plot(preds_test_mean, test_df$zorf,
     main = "Posterior Pred Means vs. Actual zorf (Test)",
     xlab = "Predicted zorf", ylab = "Actual zorf")
abline(0, 1, col = 'red', lwd = 2)
```


```{r}
rmse_test <- sqrt(mean((preds_test_mean - test_df$zorf)^2))
rmse_test
```



