---
title: "bayesion model"
author: "Krithika Natarajan"
date: "2026-04-09"
output: pdf_document
---



```{r setup, include=FALSE}
install.packages(c('tidyr', 'dplyr', 'brms', 'ggplot2'))
library(tidyr)
library(dplyr)
library(ggplot2)
library(brms)
```



## Reading in the Data

```{r}
df <- read.csv('boston_cleaned_data.csv', colClasses = c(ZipCode = 'character'))
 
meta_cols <- c('RegionID', 'SizeRank', 'ZipCode', 'RegionType',
               'StateName', 'State', 'City', 'Metro', 'CountyName','Neighborhood',
               'HasUniversity')
date_cols <- setdiff(colnames(df), meta_cols)
```

```{r}
test_group_specificity <- function(data, response, effect, groups) {
  results <- list()
  
  for (g in groups) {
    # Base model: control for group + effect
    base_formula <- as.formula(
      paste(response, "~", g, "+", effect)
    )
    
    # Interaction model: allow effect to vary by group
    interaction_formula <- as.formula(
      paste(response, "~", g, "+", effect, "*", g)
    )
    
    model_base <- lm(base_formula, data = data)
    model_interaction <- lm(interaction_formula, data = data)
    
    cmp <- anova(model_base, model_interaction)
    
    results[[g]] <- data.frame(
      group = g,
      df_added = cmp$Df[2],
      rss_drop = cmp$RSS[1] - cmp$RSS[2],
      f_value = cmp$F[2],
      p_value = cmp$`Pr(>F)`[2]
    )
  }
  
  do.call(rbind, results)
}
```


## Converting data to long form + Early Exploration
Converting Data to long form as the data follows time series format. 
```{r}
long_df <- df %>%
  pivot_longer(cols = all_of(date_cols), 
               names_to = 'date', values_to = 'zorf') %>%
  mutate(date = as.Date(gsub('\\.', '-', gsub('^X', '', date)))) %>%
  arrange(ZipCode, date) %>%
  filter(!is.na(zorf), !is.na(date)) %>%
  mutate(
    HasUniversity = as.integer(HasUniversity),
    Neighborhood  = as.factor(Neighborhood)
  )
```


### Data Checks & Validating Numbers
The purpose of this section is to ensure that our data is complete.
```{r}
summary_stats <- long_df %>%
  summarise(
    n_observations = n(),
    n_zipcodes = n_distinct(ZipCode),
    min_date = min(date, na.rm = TRUE),
    max_date = max(date, na.rm = TRUE)
  )
print(summary_stats)
```

```{r}
month_stats <- long_df %>%
  mutate(year_month = floor_date(date, unit = "month")) %>%
  summarise(
    n_unique_months = n_distinct(year_month),
    first_month = min(year_month, na.rm = TRUE),
    last_month = max(year_month, na.rm = TRUE)
  )

print(month_stats)
```

```{r}
long_df %>%
  group_by(ZipCode) %>%
  summarise(n_obs = n(), .groups = "drop") %>%
  arrange(n_obs) %>%
  print()
```
None of the zipcodes are missing data as the number of observations matches number of months represented overall

## EDA/ Feature Engineering
### Do zipcodes explain variation in rental prices?
```{r}
anova(
  lm(zorf ~ time_index, data = long_df),
  lm(zorf ~ time_index + ZipCode, data = long_df)
)
```
### Do weather patterns (seasons) impact ZORF?
```{r}
long_df <- long_df %>%
  mutate(
    month_num  = as.numeric(format(date, '%m')),
    sin_season = sin(2 * pi * month_num / 12),
    cos_season = cos(2 * pi * month_num / 12)
  )
```

```{r}
#anova checking the significance of including seasonality vs. without
lm_without <- lm(zorf ~ time_index, data = long_df)

lm_with <- lm(zorf ~ time_index + sin_season + cos_season, data = long_df)

anova(lm_without, lm_with)
```
```{r}
cor(long_df$sin_season, long_df$zorf, use = "complete.obs")
cor(long_df$cos_season, long_df$zorf, use = "complete.obs")
```
**Data Decision:** Since the joint F-test is significant but individual correlations are near zero, the seasonal terms matter only as a combined pattern, not as separate effects.

Is it different by group?
```{r}
groups <- c("HasUniversity", "Neighborhood", "ZipCode")

test_group_specificity(
  data = long_df,
  response = "zorf",
  effect = "sin_season + cos_season",
  groups = groups
)
```
The smooth sinusoidal seasonality (sin/cos) is globally consistent across locations, but it does not capture the true leasing cycle pattern.


### Are there significant differences in long-term trend over time? Are some zips appreciating faster than others over time?
```{r}
summary(lm(zorf ~ time_index, data = long_df))
```
Time is statistically signicant, which makes sense given inflation on rent prices over time. 

```{r}
groups <- c("HasUniversity", "Neighborhood", "ZipCode")

test_group_specificity(
  data = long_df,
  response = "zorf",
  effect = "time_index",
  groups = groups
)
```
The rate of appreciation over time is significantly different by zip code and by neighborhood, but is consistent between university and non-university areas (p = 0.65).

```{r}
zip_trends <- long_df %>%
  group_by(ZipCode) %>%
  do(tidy(lm(zorf ~ time_index, data = .))) %>%
  filter(term == "time_index") %>%
  select(ZipCode, estimate, std.error, p.value) %>%
  arrange(desc(estimate))
```

```{r}
ggplot(long_df, aes(x = time_index, y = zorf, 
                    group = ZipCode, color = Neighborhood)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  theme_minimal() +
  labs(title = "Zip-level trends over time (colored by Neighborhood)", 
       x = "Time Index", y = "ZORF")
```
Looking at trends in ZORF over time though, most of the lines follow a similar slope, just starting intercepts are different. There are two outliers.


So, next testing whether neighborhood labels explain a statistically significant portion of the variation in zip-level time slopes.
```{r}
# get zip-level slopes
zip_slopes <- long_df %>%
  group_by(Neighborhood, ZipCode) %>%
  summarise(
    estimate = coef(lm(zorf ~ time_index))[2],
    .groups = "drop"
  )

# Null: slopes centered around global mean
m_null <- lm(estimate ~ 1, data = zip_slopes)

# Alternative: slopes vary by neighborhood
m_neighborhood <- lm(estimate ~ Neighborhood, data = zip_slopes)

anova(m_null, m_neighborhood)
```
Yes, knowing which neighborhood a zip code belongs to tells you something meaningful about how fast that zip's rents are appreciating over time because p is less than 0.05

### Volatility
```{r}
calculate_volatility <- function(data, window = 6) {
  data %>%
    arrange(ZipCode, date) %>%
    group_by(ZipCode) %>%
    mutate(volatility = zoo::rollapply(
      zorf, 
      width   = window, 
      FUN     = sd, 
      fill    = NA, 
      align   = "right"   # only looks backwards
    )) %>%
    ungroup()
}

long_df <- calculate_volatility(long_df)
```

```{r}
# is it significant at all?
summary(lm(zorf ~ volatility, data = long_df))
```


```{r}
# is it global or group specific?
test_group_specificity(long_df %>% filter(!is.na(volatility)),
                       response = "zorf",
                       effect   = "volatility",
                       groups   = c("HasUniversity", "Neighborhood", "ZipCode"))
```



```{r}
zip_volatility_slopes <- long_df %>%
  filter(!is.na(volatility)) %>%
  group_by(Neighborhood, ZipCode) %>%
  do(tidy(lm(zorf ~ volatility, data = .))) %>%
  filter(term == "volatility") %>%
  select(Neighborhood, ZipCode, estimate) %>%
  left_join(distinct(long_df, ZipCode, HasUniversity), by = "ZipCode")

```
Neighborhood explains a statistically significant portion of the variation in how strongly volatility affects zip-level rents (F=2.83, p=0.035), justifying volatility as a neighborhood-level random slope rather than a global fixed effect.

```{r}
# does neighborhood explain zip-level slope variance?
anova(
  lm(estimate ~ 1,            data = zip_volatility_slopes),
  lm(estimate ~ Neighborhood, data = zip_volatility_slopes)
)
```

**Does neighborhood explain additional slope variance beyond what HasUniversity already captures?** 
```{r}
anova(
  lm(estimate ~ HasUniversity,                data = zip_volatility_slopes),
  lm(estimate ~ HasUniversity + Neighborhood, data = zip_volatility_slopes)
)
```
So neighborhood does not significantly explain additional variance beyond what HasUniversity already captures (p=0.076)

### Should time_index and volatility be represented dependentally? Test of collinearity.
```{r}
neighborhood_slopes <- long_df %>%
  group_by(Neighborhood) %>%
  summarise(
    time_slope = coef(lm(zorf ~ time_index))[2],
    vol_slope  = coef(lm(zorf ~ volatility))[2],
    .groups = "drop"
  )
```

```{r}
cor.test(
  neighborhood_slopes$time_slope,
  neighborhood_slopes$vol_slope
)
```
Neighborhoods where rent increases faster over time tend to have slightly higher volatility, but the relationship is extremely weak and practically negligible. 


## Train - Test Split

```{r}
# Standardize variables (mean 0, sd 1)
long_df$volatility_scaled <- as.numeric(scale(long_df$volatility))
long_df$time_index_scaled <- as.numeric(scale(long_df$time_index))
```

ensuring that the train test split is by date. 
```{r}
cutoff_date <- as.Date(quantile(long_df$date, 0.8, type = 1))

train_df <- long_df %>% filter(date <= cutoff_date)
test_df  <- long_df %>% filter(date >  cutoff_date)

cat('nrow train_df:', nrow(train_df), '\n')
cat('nrow test_df:',  nrow(test_df),  '\n')
cat('train date range:', as.character(min(train_df$date)), 
    'to', as.character(max(train_df$date)), '\n')
cat('test date range:',  as.character(min(test_df$date)),  
    'to', as.character(max(test_df$date)),  '\n')
```

## Designing The Bayesian model
I have conviction around the fact that the base rent is positive, probably around 2000K.
```{r}
# Gamma model priors
gamma_priors <- c(
  prior(normal(7.5, 0.5), class = "Intercept"),
  prior(normal(0, 0.5), class = "b", coef = "cos_season"),
  prior(normal(0.1, 0.2), class = "b", coef = "HasUniversity"),

  # zip-level random intercept SD
  prior(exponential(2), class = "sd", group = "ZipCode", coef = "Intercept"),

  # neighborhood-level random slope SDs
  prior(exponential(10), class = "sd", group = "Neighborhood", coef = "time_index_scaled"),
  prior(exponential(10), class = "sd", group = "Neighborhood", coef = "volatility_scaled"),

  # Gamma shape parameter
  prior(gamma(2, 1.5), class = "shape")
)
```

```{r}
# Fit the Gamma mixed model
gamma_brm <- brm(
  formula = zorf ~ cos_season + HasUniversity +
    (1 | ZipCode) +
    (0 + time_index_scaled + volatility_scaled | Neighborhood),
  family = Gamma(link = "log"),
  data = train_df,
  chains = 4,
  cores = getOption("mc.cores", 1),
  iter = 4000,
  warmup = 180,
  thin = 1,
  prior = gamma_priors
)
```

```{r}
# Model summary
summary(gamma_brm)
```


```{r}
bayes_R2(gamma_brm)
```

```{r}
post_preds_gamma <- posterior_predict(gamma_brm)
head(post_preds_gamma[, 1:10])
```

```{r}
# Conditional effects
plot(conditional_effects(gamma_brm, "cos_season"))
plot(conditional_effects(gamma_brm, "HasUniversity"))
plot(conditional_effects(gamma_brm, "time_index"))
plot(conditional_effects(gamma_brm, "volatility"))
```


example predictions and their confidence interval
filtering out specific zips
```{r}
selected_zips <- c("02115", "02120", "02130", "02148")

train_plot_df <- train_df %>%
  filter(ZipCode %in% selected_zips)

test_plot_df <- test_df %>%
  filter(ZipCode %in% selected_zips)

```

combining them with their CI
```{r}
post_preds_test <- posterior_predict(
  gamma_brm,
  newdata = test_plot_df
)
pred_summary <- data.frame(
  post_mean = apply(post_preds_test, 2, mean),
  lower_90  = apply(post_preds_test, 2, quantile, probs = 0.05),
  upper_90  = apply(post_preds_test, 2, quantile, probs = 0.95)
)

test_plot_df <- bind_cols(test_plot_df, pred_summary)
```

plotting the results
```{r}
ggplot() +
  # observed training data
  geom_line(
    data = train_plot_df,
    aes(x = date, y = zorf, color = ZipCode),
    linewidth = 1
  ) +
  
  # 90% credible interval on test set
  geom_ribbon(
    data = test_plot_df,
    aes(x = date, ymin = lower_90, ymax = upper_90, fill = ZipCode),
    alpha = 0.18,
    color = NA
  ) +
  
  # posterior mean on test set
  geom_line(
    data = test_plot_df,
    aes(x = date, y = post_mean, color = ZipCode),
    linewidth = 1.1,
    linetype = "dashed"
  ) +
  
  # train/test split
  geom_vline(
    xintercept = as.numeric(cutoff_date),
    linetype = "dotted",
    color = "gray40",
    linewidth = 0.8
  ) +
  
  annotate(
    "text",
    x = cutoff_date,
    y = max(long_df$zorf, na.rm = TRUE) * 0.92,
    label = "Train / Test",
    color = "gray35",
    hjust = -0.05,
    size = 5
  ) +
  
  labs(
    title = "Posterior Predictive Distributions for Selected Zip Codes",
    subtitle = "Shaded ribbon = 90% credible interval | dashed = posterior mean",
    x = NULL,
    y = "ZORI ($)",
    color = "Zip Code",
    fill = "Zip Code"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

